EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-08-20 jlamp/jlamp.tex
sez. 3

\subsection{The category of hypergraphs}

@@

		We define the \emph{hypergraph $\mathcal{G}_\Sigma$ associated with $\Sigma$} taking $O_\Sigma$ as set of hyperedges, $1$ as set of nodes and $\Delta \circ \ari_\Sigma$ as the source function and $ \Delta \coar_{\Sigma}$ as target function, where $\Delta \colon \mathbb{N}\to 1^\star$ is the isomorphism of \Cref{rem:length}.
	\end{defi}

@@

 \smallskip 
 We can interpret an arbitrary hypergaph as a \emph{typed} or \emph{colored} monoidal signature: the set of nodes provides the \emph{types} or \emph{colors}, while hyperedges can be thought of as operations, with arity and coarity provided by the source and target functions, respectively. This suggests the following definition, which will be used in Section \ref{par:tg}.

@@

\subsubsection{Two  subcategories of $\hyp$}

@@

\begin{cor}$\lmo$ is quasiadhesive.
\end{cor}

@@

Our interest for hypergraphs stems from their use as a graphical representation of algebraic terms. We thus need a way to label hyperedges with symbols taken from a signature.

\begin{exa}segnature The category $\hyps$ of \emph{algebraically labelled hypergraphs} is the slice category $\hyp/\mathcal{G}^\Sigma$.
\end{exa}

\Cref{cor:mono} and \Cref{thm:slice-functors} give us immediately an adhesivity result for $\hyp_{\Sigma}$ and a characterisation of monos in it.

\begin{prop}\label[proposition]{prop:mono} Let $\Sigma$ be an algebraic signature. Then the following facts hold
	\begin{enumerate}
		\item an arrow $(h,k)$ in $\hyp_{\Sigma}$ is a mono if and only if $h$ and $k$ are injective;
		\item $\hyp_{\Sigma}$ is an adhesive category. 
	\end{enumerate}
\end{prop}

\noindent 
\begin{minipage}[l]{.8\linewidth}	
	\begin{rem}\label[remark]{rem:label}	
			\setlength{\parindent}{1.5em}
		Let $\mathcal{H}=(E, V, s, t)$ be a hypergraph, by definition we know that $U_{\hyp}(\mathcal{G}^{\Sigma})$ is the terminal object $1$, so an arrow $\mathcal{H}\rightarrow \mathcal{G}^{\Sigma}$, is determined by a function $h\colon E_\mathcal{H}\to O_\Sigma$  making the two squares on the right commutative (cfr.~\Cref{rem:length}).

		Now, consider a coprojection $v_n\colon V^n_\mathcal{H}\to  V^\star_{\mathcal{H}}$.  By the second diagram above entails that $t_{\mathcal{H}}$ factors via the inclusion $v_1\colon V_{\mathcal{H}}\to V_\mathcal{H}^{\star}$ of words of length $1$, i.e.~all hyperedges must have targets of length $1$, that is $t_{\mathcal{H}}=v_1\circ \tau_{\mathcal{H}}$ for some $\tau_{\mathcal{H}}\colon E_{\mathcal{H}}\to V_{\mathcal{H}}$.
	\end{rem}
\end{minipage} \hfill 
\begin{minipage}[r]{.15\linewidth}
	\xymatrix@R=10pt{E_{\mathcal{H}} \ar[d]_{h}\ar[r]^{s_{\mathcal{H}}} & V^\star_{\mathcal{H}} \ar[d]^{\lgh_{V_{\mathcal{H}}}} \\  O_\Sigma \ar[r]^{\ari_\Sigma}& \mathbb{N}} \hspace{1pt}\\ \xymatrix@R=10pt{E_{\mathcal{H}} \ar[d]_{h}\ar[r]^{t_{\mathcal{H}}} & V^\star_{\mathcal{H}} \ar[d]^{\lgh_{V_{\mathcal{H}}}} \\  O_\Sigma \ar[r]^{\gamma_1}& \mathbb{N}}
\end{minipage}

$\hyp_{\Sigma}$ has a forgetful functor $U_{\Sigma}\colon \hyp_{\Sigma}\to \X$ which sends $(h,k)\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ to $U_{\hyp}(\mathcal{H}$). Now, since $U_{\hyp}(\mathcal{G}^{\Sigma})=1$ 
for every set $X$ we can define $\Delta_{\Sigma}(X)\colon \Delta_{\hyp}(X)\to \mathcal{G}^{\Sigma}$  as $(?_{O_\Sigma}, !_{X})$. It is straightforward to see that in this way we get a left adjoint to $U_\Sigma$.

\begin{prop} $U_\Sigma$
	has a left adjoint $\Delta_\Sigma$.
\end{prop}
\iffalse 
\begin{proof}Let $(h, !_{V_\mathcal{H}})\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ be an object of $\hyp_{\Sigma}$, and suppose that there exists $f\colon X\to U_{\Sigma}(\mathcal{H})$. Since, $U_{\Sigma}(\mathcal{H})=U_{\X}(\mathcal{H})$ and the identity is the unit of $\Delta_\hyp \dashv U_{\hyp}$, we get a morphism $(?_{E_{\mathcal{H}}},f)\colon \Delta_{\X}(X)\to \mathcal{H}$ of $\hyp$. And then the thesis follows since we have
	\[
	(h, !_{V_{\mathcal{{H}}}})\circ (?_{E_{\mathcal{H}}}, f)=(h\circ ?_{E_{\mathcal{H}}}, !_{V_{\mathcal{{H}}}}\circ f)=(?_{0_{\Sigma}}, !_{X})=\Delta_{\hyp}(X)\]
%and the thesis follow.   
\end{proof}

@@

We extend our graphical notation of hypergraphs to labeled ones putting the label of an hyperedge $h$ inside its corresponding square.

@@

\color{red}
TO DO - questo comprende anche la sezione successiva (?) come sottosezione + versione etichettata (anche con cose monoidali) dopo

Ipergrafi scoppiati con equivalenza/congruenza -> adesività (spero) + versione etichettata (monoidale boh)

\color{black}
\paragraph{Term Graphs}\label{par:tg}
Term graphs have been adopted as a convenient way to represent terms over a signature with an explicit sharing of sub-terms,
and as such have been a convenient tool for an efficient implementation of term rewriting~\cite{Plu:TGR-ENTCS}.
We elaborate here on the presentation given in~\cite{CastelnovoGM24}.

@@

\Cref{thm:slice-functors,prop:mono}, \Cref{lem:reg} and \Cref{prop:push} allow us to recover the following result, previously proved by direct computation in \cite[Thm.~4.2]{CorradiniG05} (see also \cite[Cor.~5.15]{CastelnovoGM24} for the details of the argument presented here).
\begin{cor}\label{cor:term}
	The category $\tg$ is quasiadhesive.
\end{cor}

\begin{rem}\label{rem:map1}
	diagramma generale dei vari ipergrafi
\end{rem}

\section{Adding equivalences to hypergraphical structures}
\label{hypereq}
The use of hypergraphs equipped with an equivalence relation over their nodes has been argued as a convenient way to express concurrency in the DPO approach to rewriting~\cite{concur2006}.
This section presents the framework by means of adhesive categories, including also its version for term graphs, as a stepping stone towards an analogous characterisation for e-graphs.

>>>>

\subsection{The category of hypergraphs}\label{sub:hyp}

@@

		We define the \emph{hypergraph $\mathcal{G}_\Sigma$ associated with $\Sigma$} taking $O_\Sigma$ as set of hyperedges, $1$ as set of nodes and $\Delta \circ \ari_\Sigma$ as the source function and $ \Delta \circ  \coar_{\Sigma}$ as target function, where $\Delta \colon \mathbb{N}\to 1^\star$ is the isomorphism of \Cref{rem:length}.
	\end{defi}

@@

 \smallskip 
 We can interpret an arbitrary hypergaph as a \emph{typed} or \emph{colored} monoidal signature: the set of nodes provides the \emph{types} or \emph{colors}, while hyperedges can be thought of as operations, with arity and coarity provided by the source and target functions, respectively. This suggests the following definition, which will be used in Section \ref{par:tg}\label{sub:hyp_alg}.

@@

\subsubsection{Two  subcategories of $\hyp$}\label{sub:sub}

@@

\begin{cor}\label{cor:lmquasi}$\lmo$ is quasiadhesive.
\end{cor}

@@

Our interest for hypergraphs stems from their use as a graphical representation of algebraic terms. As  noted in \Cref{sub:hyp_alg}, an hypergraph can be thought as a colored monoidal signature. Thus labelling correspnds to the operationn of taking the slice category over a suitable hypergraph.

\begin{defi}
	Given an hypergraph $\mathcal{G}$, the category $\hyp_\mathcal{G}$ of hypergraphs \emph{labelled by $\mathcal{G}$}  or of \emph{$\mathcal{G}$-hypergraphs} is $\hyp/\mathcal{G}$.
\end{defi}

\Cref{cor:mono} and \Cref{thm:slice-functors} give us immediately an adhesivity result for $\hyp_{\mathcal{G}}$ and a characterisation of monos in it.

\begin{prop}\label{prop:mono} Let $\mathcal{G}$ be an hypergraph. Then the following facts hold
	\begin{enumerate}
		\item an arrow $(h,k)$ in $\hyp_{\mathcal{G}}$ is a mono if and only if $h$ and $k$ are injective;
		\item $\hyp_{\mathcal{G}}$ is an adhesive category. 
	\end{enumerate}
\end{prop}

\begin{exa} A useful example is given by an algebraic signature $\Sigma$. Consider the hypregraph of \Cref{exa_3}, the category $\hyps$ of
	hypergraphs labelled by $\Sigma$ is the slice category $\hyp/\mathcal{G}^\Sigma$.
\end{exa}

\noindent 
\begin{minipage}[l]{.8\linewidth}	
	\begin{rem}	\label{rem:label}
			\setlength{\parindent}{1.5em}

			Consider an algebraic signature $\Sigma$ and an hypergraph $\mathcal{H}=(E, V, s, t)$. By definition we know that $U_{\hyp}(\mathcal{G}^{\Sigma})$ is the terminal object $1$, so an arrow $\mathcal{H}\rightarrow \mathcal{G}^{\Sigma}$, is determined by a function $h\colon E_\mathcal{H}\to O_\Sigma$  making the two squares on the right commutative (cfr.~\Cref{rem:length}).

		Now, consider a coprojection $v_n\colon V^n_\mathcal{H}\to  V^\star_{\mathcal{H}}$. Since $\Delta\circ \coar_\Sigma$ is the constant function in $1$,  the second diagram aside entails, together with \Cref{rem:len}, that $t_{\mathcal{H}}$ factors via the inclusion $v_1\colon V_{\mathcal{H}}\to V_\mathcal{H}^{\star}$ of words of length $1$, i.e $\mathcal{H}$ must be algebraic.
	\end{rem}
\end{minipage} \hfill 
\begin{minipage}[r]{.15\linewidth}
	\xymatrix@R=14pt{E_{\mathcal{H}} \ar[d]_{h}\ar[r]_{s_{\mathcal{H}}} & V^\star_{\mathcal{H}} \ar[d]^{\lgh_{V_{\mathcal{H}}}} \\  O_\Sigma \ar[r]^{\Delta\circ \ari_\Sigma}& \mathbb{N}} 
	 \xymatrix@R=14pt{E_{\mathcal{H}} \ar[d]_{h}\ar[r]_{t_{\mathcal{H}}} & V^\star_{\mathcal{H}} \ar[d]^{\lgh_{V_{\mathcal{H}}}} \\  O_\Sigma \ar[r]^{\Delta\circ \coar_\Sigma}& \mathbb{N}}
\end{minipage}

\iffalse 
\smallskip 
$\hyp_{\mathcal{G}}$ has a forgetful functor $U_{\mathcal{G}}\colon \hyp_{\mathcal{G}}\to \Set/V_{\mathcal{G}}$ which sends $(h,k)\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ to $k\colon V_\mathcal{H} \to V_{\mathcal{G}}$. Now, since $U_{\hyp}(\mathcal{G}^{\Sigma}=1$ 
for every set $X$ we can define $\Delta_{\Sigma}(X)\colon \Delta_{\hyp}(X)\to \mathcal{G}^{\Sigma}$  as $(?_{O_\Sigma}, !_{X})$. It is straightforward to see that in this way we get a left adjoint to $U_\Sigma$.

\begin{prop} $U_\Sigma$
	has a left adjoint $\Delta_\Sigma$.
\end{prop}

\begin{rem}
	contenuto...
\end{rem}
\begin{proof}Let $(h, !_{V_\mathcal{H}})\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ be an object of $\hyp_{\Sigma}$, and suppose that there exists $f\colon X\to U_{\Sigma}(\mathcal{H})$. Since, $U_{\Sigma}(\mathcal{H})=U_{\X}(\mathcal{H})$ and the identity is the unit of $\Delta_\hyp \dashv U_{\hyp}$, we get a morphism $(?_{E_{\mathcal{H}}},f)\colon \Delta_{\X}(X)\to \mathcal{H}$ of $\hyp$. And then the thesis follows since we have
	\[
	(h, !_{V_{\mathcal{{H}}}})\circ (?_{E_{\mathcal{H}}}, f)=(h\circ ?_{E_{\mathcal{H}}}, !_{V_{\mathcal{{H}}}}\circ f)=(?_{0_{\Sigma}}, !_{X})=\Delta_{\hyp}(X)\]
%and the thesis follow.   
\end{proof}

@@

\smallskip

We extend our graphical notation of hypergraphs to labelled ones putting the label of an hyperedge $h$ inside its corresponding square.

@@

As we have done in \Cref{sub:sub}, we can restrict our attention to two subcategories of labelled hypergraphs.

\begin{defi}
	Let $\mathcal{G}$ be an hypergraph and $(h,k)\colon \mathcal{H}\to \mathcal{G}$  a $\mathcal{G}$-hypergraph. We say that $(h,k)$ is a \emph{left-monogamous} (resp.~\emph{separated}) $\mathcal{G}$-hypergraph if $\mathcal{H}$ is left-monogamous (resp.~separated).

	We will denote by $\lmo_{\mathcal{G}}$ and $\shyp_{\mathcal{G}}$  the corresponding subcategories of $\hyp_{\mathcal{G}}$.
\end{defi}

We can easily realize $\lmo_{\mathcal{G}}$ and $\shyp_{\mathcal{G}}$ as comma categories.

\begin{prop}
Let $M\colon \lmo \to \hyp$ and $S\colon \shyp \to \hyp$ be the inclusion functors. Let, moreover, $\Delta_{\mathcal{G}}\colon 1\to \hyp$ be the constant functor in $\mathcal{G}$. Then:
\begin{enumerate}
	\item $\lmo_{\mathcal{G}}$ is equivalent to $\comma{M}{\Delta_\mathcal{G}}$;
	\item $\shyp_{\mathcal{G}}$ is equivalent to $\comma{S}{\Delta_\mathcal{G}}$; 
\end{enumerate}
\end{prop}

\Cref{lim} allows us to easily characterize regular monos in $\hyp_{\mathcal{G}}$.
\begin{prop}
	A morphism $(f,g)$ between $(h_1, k_1)\colon \mathcal{H}\to \mathcal{G}$ and  $(h_2, k_2)\colon \mathcal{K}\to \mathcal{G}$ is a regular mono in $\lmo_{\mathcal{G}}$ if and only if $(f,g)\colon \mathcal{H}\to \mathcal{K}$ is a regular mono in $\lmo$.
\end{prop}

In turn, by \Cref{thm:slice-functors}, Corollaries \ref{cor:eq}, \ref{cor:pusho} and \Cref{lem:regmon} we get the following.

\begin{thm} Given an hypergraph $\mathcal{G}$, the following hold true:
	\begin{enumerate}
		\item an arrow $(f,g)$  if and only if is so in $\lmo$.
		\item $\lmo_{\mathcal{G}}$ is quasiadhesive.
	\end{enumerate}
\end{thm}

\begin{rem}\label{rem:tg}
	Let $\Sigma$ be an algebraic signature. Then we know that every hypergraph labelled by $\Sigma$ must be algebraic. By \Cref{rem:elem} this means that an hypergraph $(h,k)\colon \mathcal{H}\to \mathcal{G}^\Sigma$ labelled by $\Sigma$ is left-monogamous if and only if $t_{\mathcal{H}}$ is injective. This is precisely the definition of (cyclic) \emph{term graphs} in \cite{Plu:TGR-ENTCS,CorradiniG05,CastelnovoGM24}. 
\end{rem}

\iffalse 
\color{black}
\paragraph{Term Graphs}\label{par:tg}
Term graphs have been adopted as a convenient way to represent terms over a signature with an explicit sharing of sub-terms,
and as such have been a convenient tool for an efficient implementation of term rewriting~\cite{Plu:TGR-ENTCS}.
We elaborate here on the presentation given in~\cite{CastelnovoGM24}.

@@

\Cref{thm:slice-functors,prop:mono}, \Cref{lem:reg} and \Cref{prop:push} allow us to recover the following result, previously proved by direct computation in \cite[Thm.~4.2]{CorradiniG05} (see also \cite[Cor.~5.15]{CastelnovoGM24} for the details of the argument presented here).
\begin{cor}\label{cor:term}
	The category $\tg$ is quasiadhesive.
\end{cor}
\fi 
\section{Adding equivalences to hypergraphical structures}
\label{hypereq}
The use of hypergraphs equipped with an equivalence relation over their nodes has been argued as a convenient way to express concurrency in the DPO approach to rewriting~\cite{concur2006}.
This section presents the framework by means of adhesive categories, including also its version for term graphs, as a stepping stone towards an analogous characterisation for e-graphs.
